Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-01-28 analysis/paper/paper.Rmd
working on the discussion section

Long-chain fatty acids from C20:0 to C32:0 are present in only one pottery sample (SYG-TN13-E22-2#2) with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could be also associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

@@

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The only potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. It should be noted that there are two potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to a later period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two sources [@Copley2003]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants [@Shoda2018; @Kwak2020]. In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants. There is only one potsherd (SYG-TN13-E22-2#2) have a relatively enriched δ^13^C value, indicative of a C4 plant diet for the ruminant animal.

@@

Ruminant fats in the Neolithic potsherds around ca. 4500 BP. from Guijiabao and Gaoshan could be derived from deer or horse according to the finding of faunal remains from the sites, accompanied with stone artifacts for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. However, whether horses were part of local food choices in an earlier period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones mostly concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. Although wild Asian water buffalo could be another food source due to their presence since the Pleistocene [@Zhang2016], no cattle bones were found at Guijiabao or Gaoshan in an early period. Therefore, it is more likely that deer could be the resource of ruminant fats. The analysis of faunal remains from Guijiabao is still in progress which could allow us to evaluate our findings.

check isotopic results for horse, deer

Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded over the past 4500 years.

@@

A large-scale analysis of ceramic across the region may offer insight into regional variations in food choices and the spread of agriculture.

>>>>

Long-chain fatty acids from C20:0 to C32:0 are present in a pottery sample (SYG-TN13-E22-2#2) with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could be also associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

@@

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. A few number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish  [@Shoda2018; @Kwak2020]. In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants. There is only one potsherd (SYG-TN13-E22-2#2) have a relatively enriched δ^13^C value, indicative of a C4 plant diet for the ruminant animal.

@@

Ruminant fats in the Neolithic potsherds around ca. 4500 BP. from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, with stone tools associated with hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. However, whether horses were part of local food choices in the Neolithic period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. The evidence from Guijuabao also confirms a later presence of horse remains. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value located in ranges of deer or sheep fats based on a study from England [@Evershed2002]. Although the referenced isotopic values were from modern animals and should be ideally adjusted in terms of local environments, they could still provide useful information and have been used as references in many studies []. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were parts of food sources during this period [@Zhang2016]. A detailed analysis of faunal remains from Guijiabao is still in progress that will allow evaluating our findings.

It is interesting to find a non-ruminant origin that could be C3 plants.

@@

Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded over the past 4500 years. A large-scale analysis of ceramic across the region may offer insight into regional variations in food choices and the spread of agriculture.
